R for Clinical Study Reports and Submission

### 2021-10-12 project-folder.Rmd
proofread chapter 9 - 11

In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
For an organization, it is also common to have multiple ongoing clinical trials in a clinical program.

To deliver A&R results of a clinical trial project, it is teamwork that typically requires
collaborations from both statisticians and programmers. 
In this part, let's consider how to organize a clinical trial project as an A&R leader.

@@

We recommend using the R package folder structure to organize analysis scripts 
for clinical trial development. 
Using the R package folder structure to streamline data analysis work has also been proposed before 
(@marwick2018packaging, @wuanalysis).

@@

For consistency, a well-defined folder structure with potential templates
ensures project teams organize A&R work consistently across multiple projects. 
Consistent folder structure also reduces communication costs between study team members 
and enhances the transparency of projects.

@@

> Here, the `adam/` folder contains ADaM datasets. 
> The `tlf` folder contains output TLFs in RTF format.
> We put `adam/` and `tlf/` folders within the `vignettes` folder only for illustration purposes. 
> In an actual A&R report, you may have a different location to save your input and output.

@@

```
R/
├── count_by.R
└── fmt.R
```

With a formal project, it is also important to provide proper documentation for those help functions. 
We used Roxygen2 to document functions. 
For example, the header below defines each variable in fmt_est. 
More details can be found in [Chapter 10 of the R package book](https://r-pkgs.org/man.html)

@@

Reproducibility of analysis is one of the most important aspects of regulatory deliverables. 
By using the R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we illustrate how to achieve reproducibility for R and R package versions.

@@

In the `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
So we can use `.Rprofile` to set up a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).

@@

After we open the `esubdemo` project, the code in `.Rprofile` will automatically check 
the current R version is the same as we defined in `.Rprofile`

```{r, eval = FALSE}
# set up project R version
R_version <- "4.1.1"
```

If the R version mismatch, an error message is displayed as below.

The R package version is controlled in two layers. 
Firstly, we define a snapshot date in `.Rprofile`. 
The snapshot date allows us to freeze the source code repository.

@@

If the project is initiated properly  
you shall be able to see similar messages to inform how we control R package versions.

@@

As the consistent folder is defined, it also becomes easier to create specific tools 
that fit for the analysis and reporting purpose.

Below are few potential tools that can be helpful:

>>>>

In a late-stage clinical trial, the number of A&R deliverables can easily be in the hundreds. 
For an organization, it is common to have multiple ongoing clinical trials in a clinical program.

To deliver the A&R results of a clinical trial project, it is teamwork that typically requires collaborations from both statisticians and programmers. 
In this part, let's consider how to organize a clinical trial project as an A&R leader.

@@

We recommend using the R package folder structure to organize analysis scripts for clinical trial development. 
Using the R package folder structure to streamline data analysis work has also been proposed before 
(see @marwick2018packaging, @wuanalysis).

@@

For consistency, a well-defined folder structure with potential templates
ensures project teams organize the A&R work consistently across multiple projects. 
Consistent folder structure also reduces communication costs between study team members and enhances the transparency of projects.

@@

> Undeer the `vignettes/` folder, there are `adam/` folder and `tlf` folder.
> The `adam/` folder contains ADaM datasets. 
> The `tlf` folder contains output TLFs in RTF format.
> We put `adam/` and `tlf/` folders within the `vignettes` folder only for illustration purposes. 
> In an actual A&R report, you may have a different location to save your input and output.

@@

```
R
├── count_by.R
└── fmt.R
```

With a formal project, it is also important to provide proper documentation for those help functions. 
We used Roxygen2 to document functions. 
For example, the header below defines each variable in `fmt_est`. 
More details can be found in [Chapter 10 of the R package book](https://r-pkgs.org/man.html)

@@

Reproducibility of analysis is one of the most important aspects of regulatory deliverables. 
To ensure a successful reproduce, we need a controlled R environment, including the control of R version and control of R package version.  
By using the R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we illustrate how to achieve reproducibility for R and R package versions.

@@

First, we introduce the control of R version. 
In the `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
So we can use `.Rprofile` to set up a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html). 
After we open the `esubdemo` project, the code in `.Rprofile` will automatically check the current R version is the same as we defined in `.Rprofile`.
```{r, eval = FALSE}
# set up project R version
R_version <- "4.1.1"
```
If the R version mismatch, an error message is displayed as below.

@@

Next, we introduce the control of R package version, which is controlled in two layers. 
Firstly, we define a snapshot date in `.Rprofile`. 
The snapshot date allows us to freeze the source code repository.

@@

If the project is initiated properly, you shall be able to see similar messages to inform how we control R package versions.

@@

As the consistent folder is defined, it also becomes easier to create specific tools 
that fit for the analysis and reporting purpose. Below are few potential tools that can be helpful:

### 2021-10-12 project-management.Rmd
proofread chapter 9 - 11

SDLC for A&R deliverables can be defined in four stages:

@@

In an R package development, the validation or testing is completed under the `test/` folder. 
The testthat R package can be used to streamline the validation process. 
More details of the `testthat` package for R package validation can be found in Chapter 12 of the R package book.

@@

```
tests/testthat
├── test-independent-test-tlf-01-disposition.Rmd
├── test-independent-test-tlf-02-population.Rmd
├── test-independent-test-tlf-03-baseline.Rmd
├── test-independent-test-tlf-04-efficacy.Rmd
├── test-independent-test-tlf-05-ae-summary.Rmd
├── test-independent-test-tlf-06-ae-spec.Rmd
└── test-independent-test-fmt.R
```

To validate the content of a Table, we can save the last datasets ready for table generation as a `.Rdata` file.
A validator can reproduce the TLF and compares it with the original result saved in `.Rdata` file. 
A test is passed when the results match. Customers can directly review the formatting of the TLFs by
customers by comparing them with the mock-up.

To validate a Figure, we can use [snapshot testing](https://testthat.r-lib.org/articles/snapshotting.html) strategy

@@

After completion of development and required validation of all A&R deliverables, project lead runs compliance
check for a project-specific R package similar to other R packages. Using `devtools::check()` is a
convenient way to run compliance checks or `R CMD check`. `R CMD check` is an automated check of the
contents in the R package for frequently encountered issues, before submission to CRAN. Since the project-specific
R package is not submitted to CRAN, some checks can be customized and skipped in
`devtools::check()`. The project lead should work with the study team to ensure all reported errors, warnings, and
notes by `devtools::check()` are fixed.

The project lead can also use the R package `pkgdown` to build a complete website for a project-specific R
package. The `pkgdown` website is a convenient way to run all analyses in batch and 
integrate outputs in a website.

Project-specific R functions, TLF generation programs, outputs and validation tracking information can all
be explored from the website.

For example, in the `esubdemo` project, we created the `pkgdown` website at <https://elong0527.github.io/esubdemo/>.

Many of the tasks in SDLC can be completed automatically. 
An organization can leverage CI/CD workflow to automatically enable those tasks 
such as running testing cases and creating a `pkgdown` website.

For example, in `esubdemo` project, we setup [GitHub Actions](https://github.com/elong0527/esubdemo/actions)
for these work. It can be setup by using `usethis::use_github_action()` function.

>>>>

SDLC for the A&R deliverables can be defined in four stages:

@@

In an R package development, the validation or testing is completed under the `test/` folder. 
The testthat R package can be used to streamline the validation process. 
More details of the `testthat` package for R package validation can be found in [Chapter 12 of the R package book](https://r-pkgs.org/tests.html).

@@

```
tests/testthat
├── test-independent-test-tlf-01-disposition.R
├── test-independent-test-tlf-02-population.R
├── test-independent-test-tlf-03-baseline.R
├── test-independent-test-tlf-04-efficacy.R
├── test-independent-test-tlf-05-ae-summary.R
├── test-independent-test-tlf-06-ae-spec.R
└── test-independent-test-fmt.R
```

To validate the content of a table, we can save the last datasets ready for table generation as a `.Rdata` file.
A validator can reproduce the TLF and compares it with the original result saved in `.Rdata` file. 
A test is passed when the results match. 
Customers can directly review the formatting of the TLFs by
customers by comparing them with the mock-up.

To validate a figure, we can use [snapshot testing](https://testthat.r-lib.org/articles/snapshotting.html) strategy.

@@

After completion of development and required validation of all A&R deliverables, project lead runs compliance check for a project-specific R package similar to other R packages. 
`devtools::check()` is a convenient way to run compliance checks or `R CMD check`. 
`R CMD check` is an automated check of the contents in the R package for frequently encountered issues, before submission to CRAN. 
Since the project-specific R package is not submitted to CRAN, some checks can be customized and skipped in `devtools::check()`. 
The project lead should work with the study team to ensure all reported errors, warnings, and notes by `devtools::check()` are fixed.

The project lead can also use the R package `pkgdown` to build a complete website for a project-specific R package. 
The `pkgdown` website is a convenient way to run all analyses in batch and 
integrate outputs in a website, which comprehensively covers project-specific R functions, TLF generation programs, outputs and validation tracking information, etc... 
For example, in the `esubdemo` project, we created the `pkgdown` website at <https://elong0527.github.io/esubdemo/>.

Many of the tasks in SDLC can be completed automatically. 
An organization can leverage CI/CD workflow to automatically enable those tasks 
such as running testing cases and creating a `pkgdown` website. 
For example, in `esubdemo` project, we setup [GitHub Actions](https://github.com/elong0527/esubdemo/actions)
for these work.
It can be setup by using `usethis::use_github_action()` function.
